---
title: Project Forest Modeling Document      
subtitle: Prepared by Reduce.Reuse.Grow.Inc  
format: 
  html:
    code-tools: true
    embed-resources: true
    toc: true 
    theme: flatly
    page-layout: full
editor: visual
execute:
  warning: false
  message: false

---

## Abstract   
RRG has partnered with the [Friends of the Usambara](https://usambaratravels.com/) to conduct Project Forest throughout the Tanga Administrative state in northeastern Tanzania. RRG and FOU have partnered to plant 10 million trees over ~6989 hectares.  

**In total, RRG estimates project activities can sequester ~ 698,384 Mg Co2e by the year 2066.**   

## Import Libraries

```{r}
library(tidyverse)
library(patchwork)
library(leaflet)
library(sf)
library(purrr)
library(ggtext)
setwd("~/scripts/rrg/r-scripts/iso")

```



## Ex-Ante Estimates of Biomass Removals
FOU will be planting secondary forests throughout degraded areas of Tanzania. [Robinson, et al.](https://www.nature.com/articles/s41558-025-02355-5) has published a global peer reviewed dataset of potential carbon removals for secondary forests. RRG is able to access this data through the Robinson, et al. [google earth engine app](https://ee-groa-carbon-accumulation.projects.earthengine.app/view/natural-forest-regeneration-carbon-accumulation-explorer) and through their [raw data](https://zenodo.org/records/15090826). 


To allow for a deeper uncertainty analysis, RRG used the parameter (a, k, b) .tif files provided by Robinson to get parameter estimates for the prediction curves for each pixel of the project area. These estimates were then exported through this [extraction script](./gee-extract.html) and are used to calculate estimated biomass at t with the following Chapman Richard's equation (using 1/3 as m).     

$$
y = A \left( 1 - B e^{-k t} \right)^{\frac{1}{1 - m}}
$$

where:

- $y$ = biomass (Mg C per ha at year t)
- $A$ = asymptote  
- $B$ = integration constant
- $k$ = growth rate
- $t$ = time
- $m$ = shape parameter

#### Project Maps and CR Parameters   
The map below represents the a parameters for the project region, which are the asymptotic maximum size that the system can reach under ideal conditions (max tC/ha). In the Chapman-Richards model, a corresponds to the theoretical upper limit of the growth process, meaning that as time approaches infinity, growth levels off at this value.    

```{r}
#| message: false
#| warning: false

tanzGeo <- read.csv('./csv_data/tanz-cr.csv') |> select(!1) |>
  rename(b = b1, k = b1_1, coords = .geo) |>
  mutate(coords = str_extract(coords, "(?<=\\[)[^\\]]+(?=\\])")) |>
  separate_wider_delim(coords, delim = ",", names = c("lon", "lat")) |>
  mutate(across(c(lat, lon), as.numeric))


offset <- 0.000225
pal <- colorNumeric("YlOrRd", domain = tanzGeo$a)

leaflet() |>
  addTiles() |>
  addRectangles(
    data = tanzGeo,
    lng1 = ~lon - offset,
    lat1 = ~lat - offset,
    lng2 = ~lon + offset,
    lat2 = ~lat + offset,
    color = ~pal(a),
    fillOpacity = 0.8,
    stroke = FALSE,
    popup = ~paste("a parameter value:", a)
  ) |>
  setView(
    lng = mean(tanzGeo$lon),
    lat = mean(tanzGeo$lat),
    zoom = 12 
  ) |>
  addLegend(
    pal = pal,
    values = tanzGeo$a,
    title = "a parameter"
  )


```





### Growth Curves for the Project Zone  
For these forecasts, RRG aims to make the most conservative estimates possible and will assume:   
- the minimum a value of the Robinson dataset -- 88.2   
- the maximum b value of the Robinson dataset -- 0.5058      
- the median k value of the Robinson dataset -- 0.01551      


```{r}
#| fig-width: 18
#| fig-height: 6

grid <- tanzGeo |> distinct()
t <- c(0:40)

cr <- function(a, b, k, t) {
  Biomass <- a * (1 - b * exp(-k * t)) ^ (1 / (1 - 0.33))
  return(Biomass)
}

df <- tibble(t = t, Bio = cr(t = t, a = 88.2, b = 0.5058, k = 0.01551)) 
df$MaxBio <-  cr(t = t, a = 144.8, b = 0.5058, k = 0.01551)
df$RemovalRate <- c(NA, diff(df$Bio))


p1 <- df |>
  ggplot(aes(x=t)) +
  geom_line(aes(y=Bio, color = "Minimum Biomass"), linewidth = 2, show.legend = FALSE) +
  geom_line(aes(y=MaxBio, color = "Maximum Biomass"), linewidth = 2, show.legend = FALSE) +
  geom_ribbon(aes(ymin = Bio, ymax = MaxBio), position = "identity", fill = "lightgrey") +
  theme_bw() +
  labs(title = "Estimated <span style='color: #17b3b7;'>minimum</span> and <span style='color: #f35e5a;'>maximum</span> aboveground carbon accumulation for Project Forest", color = "", y = "", 
       subtitle = "Biomass MgC ha", x="Project Lifetime") +
  theme(
    legend.position = "none",
    plot.title = element_markdown(), 
    panel.grid.minor = element_blank()
    )

p2 <- df |>
  ggplot(aes(x=t, y=RemovalRate)) +
  geom_point() +
  geom_smooth(method = "loess", linetype = "longdash", alpha = 0.6, color = "black") + 
  theme_bw() +
  labs(title = "Annual Removal Rate for Project Forest", color = "", y = "", 
       subtitle = "Annual Biomass MgC ha", x="Project Lifetime") 

p1 + p2


```

### Project Timeline Assumptions   
RRG has worked with FOU to build forecasted schedules of project activities. These assumptions can be updated given new information, but provide timelines for modeling.

- Project Time Period: **2026-2066**    
- Planting Time Period: **2026-2029**  

Below is the projected yearly planting totals (in ha).  

| Year | Total Ha Planted |
|----------|----------|
| 2026   |  1808    |
| 2027   | 1417  |
| 2028   | 1530  |
| 2029   | 2234  |
| **Total**   | **6989**  |


For the scaling the CR function, we assume that in 2026 t = 0.     

```{r}
# create function for later
growth_curve <- function(t, ha) {
  biomass <- (cr(t = t, a = 88.2, b = 0.5058, k = 0.01551) * ha * 44/12)  
  return(biomass)
}

mega <- tibble(Year = c(2024:2064),
               t = c(0:40),
          AreaPlanted = c(1808, 1417, 1530, 2234, rep(0,37)),
          Cycle1 = growth_curve(t, AreaPlanted[1]),
          Cycle2 = c(0, growth_curve(1:40, AreaPlanted[2])),
          Cycle3 = c(0, 0, growth_curve(1:39, AreaPlanted[3])),
          Cycle4 = c(rep(0,3), growth_curve(1:38, AreaPlanted[4])),
          TotalRemovals = Cycle1 + Cycle2 + Cycle3 + Cycle4)

mega_mod <- mega |>
  pivot_longer(col = Cycle1:Cycle4, names_to = "cycle", values_to = "MgC02e")

mega_mod |>
  ggplot(aes(x=Year, y = MgC02e, color = cycle)) +
  geom_point() +
  theme_bw() +
  labs(title = "Biomass Removals per Planting Cycle", y = "Mg Co2e", color = "Planting Cycle")


```

### Project Emissions      
RRG and FOU will need to collaborate to conduct a cradle-to-grave life cycle assessment (LCA) to calculate emissions for project activities in three main pools -- embodied, transportation and energy use. RRG and FOU have not conducted this analysis yet, and will employ a 20% discount rate.    

```{r}
#helper for later
discount <- function(metric, scale) {
  mass <- metric * scale
  return(mass)
}
emissions <- 0.2

```

### Leakge     
Isometric handles the leakage assessment during validation events, however it's important RRG considers leakage in removal models. For this analysis, RRG assumes a **20%** leakage rate (subject to change).   
For ex-ante estimations, leakage must be quantified in tCo2. To do this with the default 25% rate, RRG takes 25% of total Co2e removals per year.       

```{r}
leak <- 0.20

```

### Uncertainty      
RRG aims to further address uncertainty in performance through monte carlo simulation. However, for now RRG will utilize a blanket 10% uncertainty rate. In practice, RRG takes 10% of the total Co2e removals per year.    

```{r}
unc <- 0.1

```

### Performance Benchmark      
Isometric is in charge of quantifying the dynamic performance benchmark. The PB will most likely sway our estimates of MgC/ha. So, RRG will later utilize a monte carlo simulation to provide a range of performance scenarios.    

```{r}
pb <- 0

```


## Net Carbon Removals      

Simply, net removals are quantified as    

$$
Net Removals  (Mg Co2e) = (Gross Removals  (Mg Co2e) * Discount Factors) - Emissions    
$$



```{r}
#| eval: true 
#| fig-width: 18
#| fig-height: 6

mega2 <- mega |>
  mutate(NetRemovals = ((TotalRemovals * (1 - pb) * (1 - unc)) - discount(TotalRemovals, leak)) 
         - discount(TotalRemovals, emissions)) 

mega2$RemovalRate <- c(NA, diff(mega2$NetRemovals))

p1 <- mega2 |>
  ggplot(aes(x=Year, y = NetRemovals)) +
  geom_point() +
  theme_bw() +
  labs(title = "Modeled Net C02e Accumulation for Project Forest", color = "", y = "Net Mg C02e", x="Year") +
  geom_vline(xintercept=2026.5, color = "red") +
  geom_text(aes(x = 2032, y = 750000, label = "Large spike as project scales")) +
  geom_text(aes(x = 2053, y = 730000, label = paste("Total Removals:", 
                                                    format(max(mega2$NetRemovals), big.mark = ",", nsmall = 2), "Mg Co2e")))

p2 <- mega2 |>
  ggplot(aes(x=Year, y=RemovalRate)) +
  geom_bar(stat="identity") +
  theme_bw() +
  labs(title = "Annual Removal Rate for Project Forest", y = "Annual Mg C02e Removals", x="Year") +
  annotate("segment", x = 2030, xend = 2033, y = 100000, yend = 100000, colour = "red", size=0.5, alpha=0.6, arrow=arrow()) +
  geom_text(aes(x = 2039, y = 100000, label = "Large spike as project scales"),size=4) 

p1 + p2

```

### Model Uncertainty Analysis       
There are multiple assumptions that go into the net Co2e removals from project activities. To review, below is a table of the assumptions used in model calculations.     

**Assumptions**

| Model Input | Statistic |
|----------|----------|
| **Project Details**    |      |
| Project Size     | 6989 ha     |
| **Removals**    |      |
| Growth Curves    | Max tC/ha : 54.92     |
| Total Co2e Removals   | Max tCo2e: 698,384 tCo2e    |
| **Emissions**    |      |
| Estimated rate    | 20%    |
| **Discount Rates**    |      |
| Leakage    | 20%    |
| Uncertainty    | 10%    |
| Performance Benchmark    | 0% (assuming optimal performance)    |












